---
title: "First delivery Notebook How do education and social features affect STEM salaries? "
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Introduction

This notebook is the first delivery for Data Valorization course, in Université Cote de Azur, Nice (2023) . You can watch the explanation for the aim of the project and the data cleaning [here](https://www.youtube.com/watch?v=QIKk9Wc0Io89).

-   Business goal: Analyze how social features( race, gender)and education affect STEM salaries.

-   Technical goal: Perform regression to predict total yearly income

## Authors

Tymoteusz Igor Ciesielski [*tymoteusz-igor.ciesielski\@etu.univ-cotedazur.fr*](mailto:tymoteusz-igor.ciesielski@etu.univ-cotedazur.fr){.email}

Lucia Trillo Carreras [*lucia.trillo-carreras\@etu.univ-cotedazur.fr*](mailto:lucia.trillo-carreras@etu.univ-cotedazur.fr){.email}

Ewa Kupczyk [*ewa.kupczyk\@etu.univ-cotedazur.fr*](mailto:ewa.kupczyk@etu.univ-cotedazur.fr){.email}

Marina Bueno Garcia [*marina.bueno-garcia\@etu.univ-cotedazur.fr*](mailto:marina.bueno-garcia@etu.univ-cotedazur.fr){.email}

## Libraries

```{r}
library(ggplot2)
library(plotly) #interactive visualisation
library(lattice) #data exploration (corellations)
library(dplyr)
library(countrycode) # For adding continent variable
library(corrplot)
library(rpart)
library(rpart.plot)
library(randomForest)
library(caTools)
library(MLmetrics)
require(tidyr)
require(dplyr)


```

``` {library(tidyverse)}
```

## Palettes

```{r}
palette1 <- colorRampPalette(c("#803934","#3f67ab","#613fab"))
palette2 <-colorRampPalette(c("#5000ff","#d900ff","#ff0077"))
palette3 <-colorRampPalette(c("#5e9159","#599190","#645991"))

```

## Import dataset

```{r}
#df <- read.csv('C:/Users/Tymoteusz/Desktop/DataValorisation/Project/Levels_Fyi_Salary_Data.csv')
df <- read.csv('C:/Users/ltcmu/OneDrive/Documentos/Master/Data Valorisation/Proyecto/Levels_Fyi_Salary_Data.csv')
```

# Data preprocessing

In this section, we do the data preprocessing of each variable:

-   Eliminate variables that do not contain any values for all the records, or the ones that are irrelevant

```{r}
df <- subset(df, select = -c(rowNumber,
                             Some_College,
                             Highschool,
                             tag,
                             otherdetails,
                             dmaid, 
                             cityid,
                             stockgrantvalue,
                             level)
             )
```

-   Split country location into states and cities

```{r}
for (i in 1:length(df$location)) {
  aux <- strsplit(df$location[i], ", ")[[1]]
  if(length(aux) > 2){
    df$country[i] = aux[3]
  }
  else{
    df$country[i] = "US" # American records contain only city and state. Other records are splitted into 3
  }
  df$state[i] <- aux[2]
  df$city[i] <- aux[1]
}
```

-   Create a continent variable using a lookup table to match countries with their continents

```{r}
country_continent <- data.frame(
  country <- c("US", "United Kingdom", "Ireland", "India", "Belarus", "Canada", "Russia", "Netherlands", "Switzerland", "Singapore", "Germany", "Japan", "Sweden", "Australia", "United States", "Israel", "Poland", "China", "Austria", "Luxembourg", "Czech Republic", "France", "Pakistan", "New Zealand", "Denmark", "Hong Kong (SAR)", "South Africa", "Spain", "United Arab Emirates", "Hungary", "Brazil", "Bulgaria", "Philippines", "Indonesia", "Puerto Rico", "Taiwan", "Romania", "Mexico", "Costa Rica", "Marshall Islands", "Vietnam", "Panama", "Argentina", "Norway", "Moldova", "Estonia", "Kenya", "Turkey", "Italy", "Lithuania", "Nigeria", "Korea", "Ukraine", "Jordan", "Thailand", "Colombia", "Serbia", "Portugal", "Guatemala", "Yugoslavia", "Uruguay", "Slovakia", "Bangladesh", "Finland", "Chile", "Malaysia", "Latvia", "Saudi Arabia", "Peru", "Netherlands Antilles", "Belgium", "Burma", "Qatar", "Ghana", "Kazakhstan", "Uzbekistan", "Armenia", "Morocco", "Iraq", "Trinidad and Tobago", "Egypt"),
  continent <- c("North America", "Europe", "Europe", "Asia", "Europe", "North America", "Europe", "Europe", "Europe", "Asia", "Europe", "Asia", "Europe", "Oceania", "North America", "Asia", "Europe", "Asia", "Europe", "Europe", "Europe", "Europe", "Asia", "Oceania", "Europe", "Asia", "Africa", "Europe", "Asia", "Europe", "South America", "Europe", "Asia", "Asia", "North America", "Asia", "Europe", "North America", "North America", "Oceania", "Asia", "North America", "South America", "Europe", "Europe", "Europe", "Africa", "Europe", "Europe", "Europe", "Africa", "Asia", "Europe", "Asia", "Asia", "South America", "Europe", "Europe", "North America", "Europe", "South America", "Europe", "Asia", "Europe", "South America", "Asia", "Europe", "Asia", "South America", "North America", "Europe", "Asia", "Asia", "Africa", "Asia", "Asia", "Asia", "Africa", "Asia", "North America", "Africa")
)

# We add a continent variable to df, using our lookup table
df$continent <- country_continent$continent[match(df$country, country_continent$country)]
```

-   Extract the year value

```{r}
df$timestamp <- sapply(strsplit(as.character(df$timestamp), " "), `[`, 1) # First we filter out the time - we will not use it
sapply(strsplit(df$timestamp ,"/"), `[`, 3) -> df$timestamp # We retrieve the year

timestamp_proportions <- 100*table(df$timestamp)/length(df$timestamp)
sort(timestamp_proportions)
```

-   Unify names that correspond to the same company

```{r}
tolower(df$company)->df$company
df[grepl("moody", df$company),]["company"] <- "moody"
df[grepl("microsoft", df$company),]["company"] <- "microsoft"
df[grepl("amazon", df$company),]["company"] <- "amazon"
df[grepl("deloitte", df$company),]["company"] <- "deloitte"
df[grepl("verizon", df$company),]["company"] <- "verizon"
df[grepl("apple", df$company),]["company"] <- "apple"
df[grepl("genomics", df$company),]["company"] <- "genomics"
df[grepl("accenture", df$company),]["company"] <- "accenture"
df[grepl("adobe", df$company),]["company"] <- "adobe"
df[grepl("alibaba", df$company),]["company"] <- "alibaba"
df[grepl("akamai", df$company),]["company"] <- "akamai"
df[grepl("walmart", df$company),]["company"] <- "walmart"
df[grepl("jpmorgan", df$company),]["company"] <- "jpmorgan"
df[grepl("google", df$company),]["company"] <- "google"
df[grepl("facebook", df$company),]["company"] <- "facebook"
df[grepl("oracle", df$company),]["company"] <- "oracle"
df[grepl("intel", df$company),]["company"] <- "intel"
df[grepl("salesforce", df$company),]["salesforce"] <- "intel"
```

-   Select the companies with the highest frequency

```{r}
company_proportions <- 100*table(df$company)/length(df$company)
sort(company_proportions,decreasing=TRUE)[1:9]

selected_companies <- sort(company_proportions,decreasing=TRUE)[1:9]
df$company[!(df$company %in% names(selected_companies))] <- "Other" # Now we change name of all companies that do not belong to top 9 to "Other"
company_proportions <- 100*table(df$company)/length(df$company)
selected_companies <- sort(company_proportions,decreasing=TRUE)[1:10] # And include it as the 10th one 
```

-   Filter outliers of the variable totalyearlycompensation

```{r}
min(df$totalyearlycompensation)
max(df$totalyearlycompensation)
sum(is.na(df["totalyearlycompensation"])) # There should be no records with missing salary 
df <- df[df$totalyearlycompensation <= 2500000,] # We filter out people that earn more than 2.5 million $ per year as outliers
max(df$totalyearlycompensation)
```

-   Select the main titles and define the rest as "Other"

```{r}
title_proportions <- 100*table(df$title)/length(df$title)
sort(title_proportions,decreasing=TRUE)[1:5]
title_proportions <- sort(title_proportions,decreasing=TRUE)[1:5]
df$title[!(df$title %in% names(title_proportions))] <- "Other"
```

# Plots

In this section, we plot several graphs to get different insights.

-   Plots the companies that the employees were hired in

```{r}
unique(df$company)
q <- ggplot(data.frame(df$company), aes(x=df$company)) + geom_bar(fill = "steelblue")
q + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  ggtitle("Companies") + xlab("Company Name")
q
```

-   Visualization of total yearly compensation and years of experience

```{r}
d <- ggplot(df, aes(x = totalyearlycompensation, y = yearsofexperience)) + 
  geom_point()
d + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

-   Plots for countries versus companies

```{r}
b <- subset(df, country %in% c("US", "India", "United Kingdom", "Germany", "Canada"))
unique(b$country)
q <- ggplot(data.frame(b$country), aes(x=b$country)) + geom_bar()
q + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
a <- subset(df, country != "US")
c <- ggplot(a, aes(x=country)) + geom_bar()
c + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

-   Scatter plot of location against country

```{r}
d <- ggplot(a, aes(x = company, y = country)) + 
  geom_point()
d + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
df <- subset(df, select = -c(location))
```

-   Histogram with years of experience

```{r}
df <- df[df$yearsofexperience < 60,]
hist(df$yearsofexperience, main = "Years of experience", xlab="Years of experience", col=palette1(5))
```

-   Histogram with yearsatcompany

```{r}
min(df$yearsatcompany)
max(df$yearsatcompany)
sum(is.na(df$yearsatcompany))
hist(df$yearsatcompany, main = "Years at company", xlab="Years at comapny",xlim=c(0,40), col=palette3(5))
```

-   Histogram with basesalary

```{r}
min(df$basesalary)
max(df$basesalary)
summary(df$basesalary)
count(df[df$basesalary > 1000000,])

df <- df[df$basesalary <= 1000000,]
hist(df$basesalary, main = "Base salary", xlab="Base salary", col=palette2(5))
d <- ggplot(df, aes(x = basesalary, y = yearsofexperience)) + 
  geom_point()
d + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
d
```

### Eliminating NA for Race,Education and Gender

Maintaining samples with missing race,education and gender was useful for our analysis and plotting, as there. For modelling, it is necessary to eliminate

```{r}
# It occurs that we should drop the records with NA values for the following fields
df[!is.na(df$gender),]
df <- df[!is.na(df$gender),]
df <- df[!is.na(df$Education),]
df <- df[!is.na(df$Race),]
```

# One-hot encoding

The majority of our variables are categorical, so we one-hot encode them to be able to perform a deeper analysis with them (see second delivery notebook).

```{r}
names(df)

# Selecting the variables
df_onehot <-df[c("Bachelors_Degree","Masters_Degree","Doctorate_Degree","Race_Asian","Race_White","Race_Hispanic","Race_Black","Race_Two_Or_More","gender","yearsofexperience","yearsatcompany","continent","title","company","totalyearlycompensation","Education")]
#Lowercase titles
names(df_onehot)<- tolower(names(df_onehot))
names(df_onehot)

## ONE-HOT ENCODING

## Education

df_onehot<-rename(df_onehot,  ed_bachelor=bachelors_degree)
df_onehot<-rename(df_onehot, ed_master=masters_degree)
df_onehot<-rename(df_onehot,  ed_doctor=doctorate_degree)



# We will have only 4 columns for education, if someone is not, bsc, msc or phd he is other
df_onehot['ed_other'] = replicate(nrow(df_onehot), 0) # We create new column of zeros
df_onehot <- within(df_onehot, ed_other[df_onehot$ed_bachelor==0 & df_onehot$ed_master == 0 & df_onehot$ed_doctor ==0] <- 1)
# We verify if we do not have unclassified records
stopifnot(nrow(df_onehot[df_onehot$ed_bachelor==0 & 
                           df_onehot$ed_master==0 & 
                           df_onehot$ed_doctor==0 & 
                           df_onehot$ed_other==0,])== 0)

#we eliminate the variable now that it is encoded
df_onehot <-subset(df_onehot, select = -c(education) )
# Gender
# We filter out all records that do not contain valid gender
unique(df_onehot$gender)
df_onehot <- subset(df_onehot, gender !="Title: Senior Software Engineer")
unique(df_onehot$gender)
df_onehot %>% 
  mutate(gender_female = ifelse(gender=='Female', 1, 0),
         gender_male = ifelse(gender=='Male', 1, 0),
         gender_other = ifelse(gender=='Other', 1, 0),
  )  -> df_onehot

#we eliminate the variable gender now that it is encoded
df_onehot <-subset(df_onehot, select = -c(gender) )


# Company
unique(df_onehot$company)
df_onehot %>% 
  mutate(com_oracle = ifelse(company=="oracle", 1, 0),
         com_other = ifelse(company=="Other", 1, 0),
         com_amazon= ifelse(company=="amazon", 1, 0),
         com_apple = ifelse(company=="apple", 1, 0),
         com_salesforce = ifelse(company=="salesforce", 1, 0),
         com_facebook = ifelse(company=="facebook", 1, 0),
         com_google = ifelse(company=="google", 1, 0),
         com_intel = ifelse(company=="intel", 1, 0),
         com_microsoft = ifelse(company=="microsoft", 1, 0),
         com_ibm = ifelse(company=="ibm", 1, 0),
  )  -> df_onehot
stopifnot(nrow(df_onehot[df_onehot$com_amazon==0 &
                           df_onehot$com_apple==0 &
                           df_onehot$com_facebook==0 &
                           df_onehot$com_google==0 &
                           df_onehot$com_ibm==0 &
                           df_onehot$com_intel==0 &
                           df_onehot$com_microsoft==0 &
                           df_onehot$com_oracle==0 &
                           df_onehot$com_other==0 &
                           df_onehot$com_salesforce==0,])
          ==0)
#we eliminate the variable now that it is encoded
df_onehot <-subset(df_onehot, select = -c(company) )

# Titles
unique(df_onehot$title)
df_onehot %>% 
  mutate(title_pmanager = ifelse(title=="Product Manager", 1, 0),
         title_sweng = ifelse(title=="Software Engineer", 1, 0),
         title_swengman = ifelse(title=="Software Engineering Manager", 1, 0),
         title_datasci = ifelse(title=="Data Scientist", 1, 0),
         title_other = ifelse(title=="Other", 1, 0),
         title_hweng = ifelse(title=="Hardware Engineer", 1, 0),
  )  -> df_onehot
stopifnot(nrow(df_onehot[df_onehot$title_datasci==0 & 
                           df_onehot$title_hweng==0 & 
                           df_onehot$title_other==0 & 
                           df_onehot$title_pmanager==0 & 
                           df_onehot$title_sweng==0 & 
                           df_onehot$title_swengman==0,])==0)
#we eliminate the variable now that it is encoded
df_onehot <-subset(df_onehot, select = -c(title))

# Continents
unique(df_onehot$continent)
df_onehot %>% 
  mutate(continent_namerica = ifelse(continent=="North America", 1, 0),
         continent_samerica = ifelse(continent== "South America", 1, 0),
         continent_africa = ifelse(continent==  "Africa" , 1, 0),
         continent_oceania = ifelse(continent== "Oceania", 1, 0),
         continent_asia = ifelse(continent=="Asia", 1, 0),
         continent_europe = ifelse(continent=="Europe", 1, 0),
  )  -> df_onehot

#we eliminate the variable now that it is encoded
df_onehot <-subset(df_onehot, select = -c(continent) )

```

## Exporting the dataset

```{r}
names(df_onehot)

write.csv(df_onehot, '/Users/ltcmu/OneDrive/Documentos/Master/Data Valorisation/Proyecto/clean_salary_data.csv', row.names=FALSE)

```
